Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-12-22 paper.Rnw
Experiment 5 write up overhauled. WAval bland-altman plot

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
%  - remove 'accuracy' unless specifically talking relative to ground truth (and
%    even then, qualify)
%  - overlap/similarity --> agreement? choose one. 
%  - check that chart labels are sensible 
%  - in particular, check that Experiment 5 match Experiment 1
%  - fix figure 2 schematic language for the experiments (not 2a 2b). 
% 
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))
lm_eqn = function(m) {
  l <- list(a = format(coef(m)[1], digits = 2),
      b = format(abs(coef(m)[2]), digits = 2),
      r2 = format(summary(m)$r.squared, digits = 3));
  if (coef(m)[2] >= 0)  {
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
  } else {
    eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l)    
  }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}
<<WAval-prep, cache=T>>=
wa.xval.data = subset(read.csv('data/WAval.csv'))
wa.xval.casted = dcast(melt(wa.xval.data, id.vars=c("file","version")), file + variable ~ version)
wa.xval.casted = rename(wa.xval.casted, c("variable"="region"))
levels(wa.xval.casted$region) <-list("CA1"="X1", "CA1"="X101", 
                            "Subiculum"="X2","Subiculum"="X102", 
                            "CA4/DG"="X4","CA4/DG"="X104", 
                            "CA2/CA3"="X5","CA2/CA3"="X105", 
                            "SR/SL/SM"="X6","SR/SL/SM"="X106")
@
This experiment explores \mb{} segmentation of hippocampal subfields. To achieve
this, a leave-one-out validation was conducted in which lower-resolution images
($0.9mm^3$ voxels) of each Winterburn atlas subject were segmented using the
remaining high-resolution Winterburn atlas subjects' images.

@@

ggplot(wa.xval.volumes, aes(x=gold_0.3mm, y=volume, colour=measure)) + 
  facet_wrap(~region, ncol=3, scales='free') + 
  geom_point() + 
  coord_equal() + 
  geom_smooth(method='lm')  + 
  xlab(expression("Manual Unilateral Volume" (mm^3))) +
  ylab(expression("Automated Unilateral Volume" (mm^3))) +
  geom_text(aes(x=xoffset, y=yoffset, label=eqn), parse=TRUE,
    data=wa.xval.vols.lm, size=2,hjust=0) +
  theme(legend.direction = "horizontal", legend.position = "bottom")

@@

@@

\mb{}-based segmentations using
both \animal{} and \ants{} registration algorithms demonstrate good agreement with \snt{}
segmentations (maximum mean DSC of \Sexpr{round(snt.mb.kappa.best, digits=2)}
when using 9 atlases, 19 templates, ANTS registration, and majority vote label
fusion); Figure \ref{fig:ADNI1-xval-k-mean}).  Qualitatively, both \animal{} and
\ants{}-based segmentations demonstrate trend overlap accuracy that increases with
the size of atlas library and template library. Improvement in accuracy plateaus
with template libraries larger than ten images.

The use of \mb{} with \ants{} registration shows a pronounced increase
in segmentation accuracy over \mb{} with \animal{} registration, across all
other variable settings we tested. Additionally, by itself, using a weighted
voting strategy did not significantly improve segmentation accuracy, contrary to
the findings of \citet{Aljabar2009} in basic \multiatlas{} segmentation. Given
these findings, in the remainder of this section only results using the \ants{}
registration algorithm and majority vote fusion will be shown.

With at least five templates, \mb{} shows a mean improvement in similarity score
over \multiatlas{}-based segmentation for the same size of atlas library with
majority vote label fusion (Figure \ref{fig:ADNI1-SNT-xval-k-diff}). The
magnitude of improvement increases with the size of template library, and shows
diminishing returns with larger atlas libraries. Peak improvement
(+\Sexpr{round(snt.mb.diff.best, digits=3)} DSC) is found with a single atlas
and template library of 19 images.

In addition to a mean increase in similarity score over \multiatlas{}-based
segmentation, \mb{} also shows more consistency in similarity scores across all
subjects and validation folds (Figure \ref{ADNI1-SNT-xval-variability}). A
template library of at least 13 images is sufficient to show significant ($p <
0.05$) decrease in variance for all sizes of atlas library tested (1-9 images).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<fig:ADNI1-SNT-xval-k-mean, cache=TRUE, dependson='ADNI1-xval-load', fig.width=5, fig.height=5>>=
stderr <- function(x) sqrt(var(x)/length(x))
ggplot(subset(snt_xval_mean, (templates.mb*atlases) %% 2 == 1),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(.~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,
               fun.ymin=function (x) mean(x) - stderr(x),
               fun.ymax=function (x) mean(x) + stderr(x), 
               geom='errorbar',
               aes(y=k.mb, weight=1, group=as.factor(atlases)), width=0.5) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_y_continuous(breaks=seq(0,1,0.05)) + 
  scale_colour_hue(name="Number of Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean similarity (DSC)" ) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text = element_text(size = 8))
@
  \caption{Mean Dice's Similarity Coefficient of \mb{} segmentations relative to
  \snt{} segmentations for 69 ADNI subjects vs. atlas and template library
  size, registration algorithm, and label fusion method. 
  Error bars indicate standard error.
  \label{fig:ADNI1-SNT-xval-k-mean}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
%  - remove 'accuracy' unless specifically talking relative to ground truth (and
%    even then, qualify)
%  - overlap/similarity --> agreement? choose one. 
%  - check that chart labels are sensible 
%  - in particular, check that Experiment 5 match Experiment 1
%  - fix figure 2 schematic language for the experiments (not 2a 2b). 
% 
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))
lm_eqn = function(m) {
  l <- list(a = format(coef(m)[1], digits = 2),
      b = format(abs(coef(m)[2]), digits = 2),
      pearson = format(sqrt(summary(m)$r.squared), digits = 3));
  if (coef(m)[2] >= 0)  {
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)~"="~pearson,l)
  } else {
    eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)~"="~pearson,l)    
  }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%    RESULTS: Experiment 2: Hippocampal Subfield Cross-Validation     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment 2 Results: Winterburn Atlases Cross-Validation}
<<WAval-prep, cache=T>>=
wa.xval.vols = subset(read.csv('data/WAval.csv'))
wa.xval.kappa = subset(read.csv('data/WAval-kappa.csv'))
wa.xval.casted = dcast(melt(wa.xval.vols, id.vars=c("file","version")), file + variable ~ version)
wa.xval.casted = rename(wa.xval.casted, c("variable"="region"))
levels(wa.xval.casted$region) <-list("CA1"="X1", "CA1"="X101", 
                            "Subiculum"="X2","Subiculum"="X102", 
                            "CA4/DG"="X4","CA4/DG"="X104", 
                            "CA2/CA3"="X5","CA2/CA3"="X105", 
                            "SR/SL/SM"="X6","SR/SL/SM"="X106")
@
This experiment explores \mb{} segmentation of hippocampal subfields. To achieve
this, a leave-one-out validation was conducted in which lower-resolution images
($0.9mm^3$ voxels) of each Winterburn atlas subject were segmented using the
remaining high-resolution Winterburn atlas subjects' images.

@@

ggplot(wa.xval.volumes, aes(x=gold_0.3mm, y=volume, colour=measure)) + 
  facet_wrap(~region, ncol=3, scales='free') + 
  geom_point() + 
  coord_equal() + 
  geom_smooth(method='lm')  + 
  xlab(expression("Manual Unilateral Volume" (mm^3))) +
  ylab(expression("Automated Unilateral Volume" (mm^3))) +
  geom_text(aes(x=xoffset, y=yoffset, label=eqn), parse=TRUE,
    data=wa.xval.vols.lm, size=2,hjust=0) +
  scale_color_discrete(name="") + 
  theme(legend.direction = "horizontal", legend.position = "bottom")

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Winterburn-XVAL: bland-altman 
\begin{figure}[h!]
<<WAval-bland,cache=T,fig.width=7,fig.height=6>>=
wa.xval.volumes = melt(wa.xval.casted, id.vars=c('file', 'region', 'gold_0.3mm'))
wa.xval.volumes = rename(wa.xval.volumes, c("variable"="measure", "value"="volume"))
levels(wa.xval.volumes$measure) <- list("Subsampled atlas labels"="gold_0.9mm",
                               "MAGeT on subsampled atlas images" = "mb_0.9mm",
                               "MAGeT on T1 BRAVO images"="mb_bravo")

wa.xval.volumes$diff = wa.xval.volumes$gold_0.3mm - wa.xval.volumes$volume 
wa.xval.volumes$mean = (wa.xval.volumes$gold_0.3mm + wa.xval.volumes$volume)/2
wa.xval.bland.limits = ddply(wa.xval.volumes,c("region"),function(df) {
  data.frame(y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T))
})

ggplot(wa.xval.volumes, aes(x=mean, y=diff, colour=measure)) + 
  facet_wrap(~region, ncol=3, scales='free') + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=1) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=wa.xval.bland.limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", 
             data = wa.xval.bland.limits, hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of manual and automated HC volume ' (mm^3))) + 
  ylab(expression('Difference of manaul and automated HC volume ' (mm^3))) + 
  scale_color_discrete(name="") + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))

@
  \label{fig:WAval-bland}
\end{figure}

@@

As when comparing against manual labels in Experiment 1, we find similar
behaviour when comparing \mb{} segmentations to \snt{} labels: similarity scores
increase with increasing numbers of atlases and templates, with diminishing
increases in improvement trending towards a plateau (Figure
\ref{fig:ADNI1-xval-k-mean}).  As in Experiment 1, using \ants{} registration
leads to significantly increased similarity scores, and there is no significant
difference in scores from any of the label fusion methods.  Mean DSC score peaks
at \Sexpr{round(snt.mb.kappa.best, digits=3)} when using 9 atlases, 19
templates, ANTS registration, and majority vote label fusion).  Compared to
\multiatlas segmentations, we find \mb{} segmentations show increasing
improvement with larger atlas and template libraries when using more than 9
templates and 5 or fewer atlases (Figure \ref{fig:ADNI1-SNT-xval-k-diff}). Peak
improvement (+\Sexpr{round(snt.mb.diff.best, digits=3)} DSC) is found with a
single atlas and template library of 19 images.  In addition to a mean increase
in similarity score over \multiatlas{}-based segmentation, \mb{} also shows more
consistency in similarity scores across all subjects and validation folds
(Figure \ref{ADNI1-SNT-xval-variability}) with a large enough template library.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<fig:ADNI1-SNT-xval-k-mean, cache=TRUE, dependson='ADNI1-xval-load', fig.width=5, fig.height=5>>=
stderr <- function(x) sqrt(var(x)/length(x))
ggplot(subset(snt_xval_mean, (templates.mb*atlases) %% 2 == 1),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases), linetype=reg_method)) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(.~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=interaction(reg_method,atlases))) + 
  stat_summary(fun.y=mean,
               fun.ymin=function (x) mean(x) - stderr(x),
               fun.ymax=function (x) mean(x) + stderr(x), 
               geom='errorbar',
               aes(y=k.mb, weight=1, group=interaction(reg_method,atlases)), width=0.5) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_y_continuous(breaks=seq(0,1,0.05)) + 
  scale_colour_hue(name="Number of Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean similarity (DSC)" ) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text = element_text(size = 8))
@
  \caption{Mean Dice's Similarity Coefficient of \mb{} segmentations relative to
  \snt{} segmentations for 69 ADNI subjects vs. atlas and template library
  size, registration algorithm, and label fusion method. 
  Error bars indicate standard error.
  \label{fig:ADNI1-SNT-xval-k-mean}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
